---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

## nls.multstart

Robust and reproducible non-linear regression in R

### Issues and suggestions

Please report any issues/suggestions for improvement in the [issues link](https://github.com/padpadpadpad/nls.multstart/issues) for the repository. Or please email d.padfield@exeter.ac.uk.

[![Travis-CI Build Status](https://travis-ci.org/padpadpadpad/nls.multstart.svg?branch=master)](https://travis-ci.org/padpadpadpad/nls.multstart)

### Licensing

This package is licensed under GPL-3.

### Overview

__nls.multstart__ is an R package that allows more robust and reproducible non-linear regression compared to __nls__ or __nlsLM__. These functions allow only a single starting value, meaning that it can be hard to get the best estimated model. This is especially true if the same model is fitted over the levels of a factor, which may have the same shape of curve, but be much different in terms of parameter estimates.

__nls_multstart()__ is the main (currently only) function of __nls.multstart__. Similar to the R package __nls2__, it allows multiple starting values for each parameter and then iterates through multiple starting values, attempting a  fit with each set of start parameters. The best model is then picked on AIC score. This results in a more reproducible and reliable method of fitting non-linear least squares regression in R.

This package is designed to work with the __tidyverse__, harnessing the functions within __broom__, __tidyr__, __dplyr__ and __purrr__ to extract estimates and plot things easily with __ggplot2__. A slightly less tidy-friendly implementation is [__nlsLoop__](https://github.com/padpadpadpad/nlsLoop).

### Installation and examples

#### 1. Installation

R packages in GitHub can be installed using __devtools__.

```{r install, message=FALSE, eval = FALSE}
# install package
devtools::install_github("padpadpadpad/nls.multstart")

```

#### 2. Run nls_multstart()

__nls_multstart()__ can be used to do non-linear regression on a single curve

```{r run_nlsLoop_prep, message=FALSE, warning=FALSE, results='hide'}

# load in nlsLoop and other packages
library(nls.multstart)
library(ggplot2)
library(broom)
library(purrr)
library(dplyr)
library(tidyr)

# load in example data set
data("Chlorella_TRC")

# define the Sharpe-Schoolfield equation
schoolfield_high <- function(lnc, E, Eh, Th, temp, Tc) {
  Tc <- 273.15 + Tc
  k <- 8.62e-5
  boltzmann.term <- lnc + log(exp(E/k*(1/Tc - 1/temp)))
  inactivation.term <- log(1/(1 + exp(Eh/k*(1/Th - 1/temp))))
  return(boltzmann.term + inactivation.term)
  }
```

```{r run_nls_mulstart}
 # subset dataset
d_1 <- subset(Chlorella_TRC, curve_id == 1)

# run nls_multstart
fit <- nls_multstart(ln.rate ~ schoolfield_high(lnc, E, Eh, Th, temp = K, Tc = 20),
                     data = d_1,
                     iter = 500,
                     param_bds = c(-10, 10, 0.1, 2, 0.5, 5, 285, 330),
                     supp_errors = 'Y',
                     AICc = 'Y',
                     na.action = na.omit,
                     lower = c(lnc = -10, E = 0, Eh = 0, Th = 0))

fit
```

#### 3. Clean up fit

This fit can then be tidied up in various ways using the R package __broom__. Each different function in __broom__ returns a different set of information. __tidy()__ returns the estimated parameters, __augment()__ returns the predictions and __glance()__ returns information about the model such as AIC score.

```{r look_at_data}
# get info
info <- glance(fit)
info

# get params
params <- tidy(fit)
params

# get predictions
preds <- augment(fit)
preds

```

#### 4. Plot fit

The predictions can then easily be plotted alongside the actual data.

```{r plot_one_fit, fig.height=6, fig.width=7}

ggplot() +
  geom_point(aes(K, ln.rate), d_1) +
  geom_line(aes(K, .fitted), preds)

```

#### 5. Fitting over levels of a factor with nls_multstart

__nls_multstart()__ is unlikely to speed you up very much if only one curve is fitted. However, if you have 10, 60 or 100s of curves to fit, it makes sense that at least some of them may not fit with the same starting parameters, no matter how many iterations it is run for. 

This is where __nls_multstart()__ can help. Multiple models can be fitted using __purrr__, __dplyr__ and __tidyr__. These fits can then be tidied using __broom__, an approach Hadley Wickham has previously [written about](https://blog.rstudio.com/2016/02/02/tidyr-0-4-0/).

```{r multiple_models}
# fit over each set of groupings
fits <- Chlorella_TRC %>%
  group_by(., flux, growth.temp, process, curve_id) %>%
  nest() %>%
  mutate(fit = purrr::map(data, ~ nls_multstart(ln.rate ~ schoolfield_high(lnc, E, Eh, Th, temp = K, Tc = 20),
                                   data = .x,
                                   iter = 1000,
                                   param_bds = c(-1000, 1000, 0.1, 2, 0.5, 10, 285, 330),
                                   supp_errors = 'Y',
                                   AICc = 'Y',
                                   na.action = na.omit,
                                   lower = c(lnc = -10, E = 0, Eh = 0, Th = 0))))
```

A single fit can check to make sure it looks ok.

```{r, check_fit}
# look at output object
fits

# look at a single fit
summary(fits$fit[[1]])
```

#### 6. Clean up multiple fits

These fits can be cleaned up in a similar way to the single fit, but this time __purrr::map()__ iterates the __broom__ function over the grouping variables.

```{r, tidy_many_fits}
# get summary
info <- fits %>%
  unnest(fit %>% map(glance))

# get params
params <- fits %>%
  unnest(fit %>% map(tidy))

# get predictions
preds <- fits %>%
  unnest(fit %>% map(augment))
```

Looking at __info__ allows us to see if all the models converged.

```{r, info}
info
```

#### 7. Plotting predictions

When plotting non-linear fits, it often looks better to have a smooth curve, even if there are not many points underlying the fit. This can be achieved by including `newdata` in the __augment()__ function and creating a higher resolution set of predictor values. 

However, when predicting for many different fits, it is not certain that each curve has the same range of predictor variables. Consequently, we need to filter each new prediction by the __min()__ and __max()__ of the predictor variables.

```{r, create_new_preds}
# new data frame of predictions
new_preds <- Chlorella_TRC %>%
  do(., data.frame(K = seq(min(.$K), max(.$K), length.out = 150), stringsAsFactors = FALSE))

# max and min for each curve
max_min <- group_by(Chlorella_TRC, curve_id) %>%
  summarise(., min_K = min(K), max_K = max(K)) %>%
  ungroup()

# create new predictions
preds2 <- fits %>%
  unnest(fit %>% map(augment, newdata = new_preds)) %>%
  merge(., max_min, by = 'curve_id') %>%
  group_by(., curve_id) %>%
  filter(., K > unique(min_K) & K < unique(max_K)) %>%
  rename(., ln.rate = .fitted) %>%
  ungroup()

```

These can then be plotted using __ggplot2__.

```{r, plot_many_fits, fig.height=7,  fig.width = 7}
# plot
ggplot() +
  geom_point(aes(K - 273.15, ln.rate, col = flux), size = 2, Chlorella_TRC) +
  geom_line(aes(K - 273.15, ln.rate, col = flux, group = curve_id), alpha = 0.5, preds2) +
  facet_wrap(~ growth.temp + process, labeller = labeller(.multi_line = FALSE)) +
  scale_colour_manual(values = c('green4', 'black')) +
  theme_bw(base_size = 12, base_family = 'Helvetica') +
  ylab('log Metabolic rate') +
  xlab('Assay temperature (ºC)') +
  theme(legend.position = c(0.9, 0.15))
```

